---
title: "BLiP for point source detection"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blipr-pointsource}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates how to use ``blipr`` to perform resolution-adaptive point source detection.

## Problem setting

Suppose we observe photon counts from a telescope, and we are interested in detecting point-sources (e.g., stars) from the image. Since most images have some blur, we may not be able to perfectly localize each point-source, but BLiP can still help localize them as precisely as possible. Below, we give an example of how to apply BLiP on top of outputs from a pretrained [StarNet model](https://github.com/Runjing-Liu120/DeblendingStarfields). Note that the StarNet model is described in detail in [Liu et. al (2021)](https://arxiv.org/pdf/2102.02409.pdf).

To start, we load simulated $20x20$ image data, which was generated using code from [Liu et. al (2021)](https://arxiv.org/pdf/2102.02409.pdf).

```{r plot_image, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(blipr)

# Load simulated image data
npixels = 20
image <- blipr::starnet_sim_data$image
true_locs <- blipr::starnet_sim_data$true_locs

# plot
plot_starnet_sim <- function() {
  raw_df <- data.frame(
    y=rep(1:npixels, each=npixels),
    x=rep(1:npixels, npixels),
    log_intensity=log(as.numeric(t(image[2,,])))
  )
  loc_df <- data.frame(
    y=npixels*true_locs[,1],
    x=npixels*true_locs[,2]
  )
  g <- ggplot() + 
    geom_tile(
      data=raw_df, 
      mapping=aes(x=x, y=y, fill=log_intensity)
    ) +
    scale_fill_gradient(low="white", high="black") +
    geom_point(
      data=loc_df, mapping=aes(x=x, y=y), 
      color='red', shape='x', size=3.5
    )
  return(g)
}
print(plot_starnet_sim())
  
```

## Applying BLiP

Next, we load the posterior samples from the pretrained StarNet model. See [here](https://github.com/Runjing-Liu120/DeblendingStarfields) or [here](https://github.com/amspector100/DeblendingStarfields) for examples of how to fit pretrained StarNet models.

Note that the posterior samples are an $(N, n_{\mathrm{source}}, d)$-shaped array, where $N$ is the number of posterior samples and $d=2$ is the dimension of the image. If ``post_samples[i, j] = (x, y)``, this means that the $i$th posterior sample has asserted that there is a point-source at location $(x,y)$. Since there may be different numbers of point-sources per iteration, NaNs are ignored. See API reference for more details on the data input format.

```{r post_samples1}
post_samples <- blipr::starnet_sim_data$post_samples
# Print the location of sources according to the first posterior sample.
# The NaN means there are only six estimated sources.
post_samples[1,,]
```

```{r post_samples2}
# In contrast, in the 8th iteration, there were five estimated sources.
post_samples[8]
```

Given these posterior samples, applying BLiP is as easy as calling the ``BLiP_cts`` function. Note that ``BLiP_cts`` is useful in problems where the set of possible signal locations is continuous, for example in this problem, where a point-source could take any location in $[0,20]^2$.

```{r}
grid_sizes <- exp(log(10)*seq(log10(4), 2.5, length.out=25))
detections <- blipr::BLiP_cts(
  post_samples,
  extra_centers=NULL, # option to add extra centers for candidate groups
  grid_sizes=grid_sizes,
  verbose=F
)
```
We can plot these detections on top of the image to see the detections.

```{r}
g <- plot_starnet_sim()
detection_df <- data.frame(
  y=npixels * sapply(detections, function(x) {x$center[1]}),
  x=npixels * sapply(detections, function(x) {x$center[2]}),
  radius=npixels * sapply(detections, function(x) {x$radius})
)
library(ggforce)
g + ggforce::geom_circle(
  aes(x0=x, y0=y, r=radius), 
  data=detection_df,
  color='cornflowerblue',
)
```

The BLiP detections are the blue circles, and as we can see, BLiP makes true detections at various resolutions based on the posterior samples.
