---
title: "BLiP for change-point detection"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{blipr-changepoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette illustrates how to use ``blipr`` to perform resolution-adaptive change-point detection.

## Synthetic data

Given time series data $(Y_1, \dots, Y_n)$, suppose we are interested in looking for "change-points", or times where the stochastic process changes. Often, we can tell that a process has changed, but we cannot discern exactly where it has changed because each observation $\{Y_i\}$ is noisy. The following synthetic dataset gives one example of this:

```{r data, warning=FALSE}
library(tidyverse)
library(blipr)

# Generate synthetic data
set.seed(1234)
n <- 200 
data <- blipr::generate_changepoint_data(
  n=n, sparsity=0.04, tau2=4
)
df <- data.frame(
  Time=c(1:n),
  Y=data$y, # observed data
  beta=data$beta, # change points
  mu=data$mu # mean of Y
)

# Plot data
ggplot(df, aes(x=Time)) + 
  geom_point(aes(y=Y, color='Y')) +
  geom_line(aes(y=mu, color='E[Y]')) + 
  scale_color_manual(values=c("Y"='blue', 'E[Y]'='red'))
```

To detect regions where the true mean of $Y$ has changed, we first use the ``bcp`` package to fit a Bayesian change-point detection model. Then, we can apply BLiP directly on top of the posterior samples.

```{r}
# Fit Bayesian change point model, running 10 chains of 10K samples
library(bcp)
post_samples <- matrix(nrow=0, ncol=n)
N <- 2000; burnin <- N/10; chains <- 10
for (i in 1:chains) {
  bcp_out <- bcp(y=data$y, return.mcmc=T, mcmc=N, burnin=burnin, p0=0.1)
  post_samples <- rbind(
    post_samples, t(bcp_out$mcmc.rhos[,(burnin+1):(N+burnin)])
  )
}
# Shift by one to ensure indexing conventions line up
post_samples <- cbind(
  matrix(0, dim(post_samples)[1], 1), 
  post_samples[,0:(n-1)]
)

# Create candidate groups and apply BLiP
cand_groups <- sequential_groups(samples=post_samples, max_pep=0.5)
detections <- blipr::BLiP(cand_groups=cand_groups)
```

We can plot these detections and check if they are true or false positives.

```{r}
# True change-points
cps <- which(df$beta != 0)
# Process detections
detection_df <- data.frame(
  start=sapply(detections, function(x) {min(x$group)}),
  end=sapply(detections, function(x) {max(x$group)}),
  true_disc=sapply(detections, function(x) {
    length(intersect(x$group, cps)) > 0
  }
)) %>% mutate(true_disc=ifelse(true_disc, "True", "False"))
# Plot
ggplot(df) + 
  geom_point(aes(x=Time, y=Y, color='Y')) +
  geom_line(aes(x=Time, y=mu, color='E[Y]')) + 
  geom_rect(data=detection_df, mapping=aes(
      xmin=start, xmax=end + 1,
      ymin=min(df$Y), ymax=max(df$Y),
      fill=true_disc, alpha=1/(2*(end-start+1)),
    ),
  ) +
  scale_color_manual(values=c("Y"='blue', 'E[Y]'='red')) + 
  scale_fill_manual(values=c("True"="forestgreen", "False"="orange")) +
  labs(fill='True Discovery') +
  scale_alpha(guide='none')
```

As shown above, BLiP has detected change-points in each of the highlighted regions. In this case, it turns out that BCP + BLiP only made true discoveries.
